The bhsdtr package: flexible bayesian inference for hierarchical or non-hierarchical equal variance normal Signal Detection Theory models with or without ratings.

### 2017-11-03 inst/preprint/luxsdt.tex
strona tytulowa i wyciete komentarze

% TODO Bardziej szczegółowo przedstawić wyniku testów na symulacji

% TODO Wspomnieć o gamma_to_crit

% TODO wyjaśnić sens prioru ln(100) dla gamma

% TODO opisać domyślne priory dla zeta

% mixing properties?

% constrained czy restricted?

\documentclass[oneside,a4paper]{article}
% \usepackage[MeX]{polski} %pakiety m.in. umozliwia rozne ciekawe rzeczy
% \usepackage[latin2]{inputenc}
% Bedzie osobnym punktem, na osobnej stronie
\usepackage[sectionbib,bibnewpage]{apacite}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{float}
\usepackage{graphicx}
\usepackage{listings}

@@

\begin{document}
\tableofcontents{}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{plot_sdtr.pdf}
  \caption{Equal variance normal Signal Detection Theory model with
    additional criteria}
  \label{fig:1}
\end{figure}
% Czułość i kryterium

@@

A not so uncommon practice of analysing data aggregated over subjects,
items or other sampled factors represents an extreme case of ignoring
hierarchical data structure. Invalidity of this approach in the
context of SDT was clearly illustrated with results of simulational
studies by \citeA{morey2008problematic}, although strictly speaking it
requires such demonstrations only in order to make a point about a
specific case. SDT is a non-linear model, therefore, by definition
expected values of estimates of its parameters based on data
aggregated over sampled factors are not in general true population
averages. Such estimates are biased and inference about the target
population based on such estimates is simply not valid. To give a
concrete example, concider two unbiased participants, one with
$d' = 2$ and one with $d' = 4$. Their expected average accuracy is
given by $(\Phi(1) + \Phi(2)) / 2 = .91$, which corresponds to
$d' = 2.68$, whereas their true average $d'$ is $3$.

@@

In the \texttt{bhsdtr} package sensitivities and criteria are derived
from unconstrained parameters by convenient isomorphisms. This is
easily achieved in case of sensitivity by deriving $d'$ from
$\delta = \ln(d')$ and modelling random effects on $d'$, if any, by
assuming that $\delta$ is normally distributed. Difficulties in
representing the hierarchical structure of the criteria stem from the
fact that even though each criterion considered in isolation is an
unconstrained parameter, the vector of all the criteria is ordered.
% The solution used in the \texttt{bhsdtr} package is based on
% the way ordered categorical variables with two or more categories,
% such as confidence ratings, are modelled in ordered logistic
% regression.
We propose to map the $R^{K-1}$ space, representing unconstrained
criteria, to the $K$ dimensional probability simplex space using the
softmax function and to map the simplex space to the ordered criteria
space by means of the inverse normal CDF. Specifically, assume that a
normal distribution is centered on the midpoint between the two
evidence distributions. The value of criterion $i$ is then given by:

@@

The mapping in Eq.~\ref{eq:1} is an isomorphism between the $R^{K-1}$
space and the space of $K-1$ ordered criteria vectors. Elements
$i = 1...K$ of the $\bm{\gamma}$ vector represent relative distances
between adjacent criteria because their exponents represent relative
magnitudes of areas under the normal curve delineated by the pairs of
adjacent criteria:
$e^{\bm{\gamma}_i} / e^{\bm{\gamma}_j} = (\Phi(\bm{c}_i) -
\Phi(\bm{c}_{i-1})) / (\Phi(\bm{c}_j) - \Phi(\bm{c}_{j-1}))$.
% Other things being equal, the greater the $\bm{\gamma}_i$ value the
% more the $\bm{c}_{j \geq i}$ criteria are shifted to the right.
When $K=2$ only $\bm{\gamma}_1$ is free to vary and its value directly
represents the direction and magnitude of bias: $\bm{\gamma}_1$ is $0$
when the criterion is placed at the midpoint between the means of
evidence distributions, and the more negative (positive) it is the
more the criterion is placed to the left (right) of midpoint. Our
experience indicates that it is often a good idea to multiply all the
criteria by a value greater than $1$ which is equivalent to making the
mapping distribution in Eq.~\ref{eq:1} wider. This tends to even out
the values of $\bm{\gamma}$, especially when criteria are widely
spread as can happen for moderate to large $d'$ values. This feature
is implemented in the \texttt{bhsdtr} package by introducing a scaling
factor.

@@

% Instalacja

@@

% gabor 13

@@

\subsubsection{Step 1: preparing the data}
% Combined response

@@

% aggregate_responses

@@

\subsubsection{Step 2: creating the model code}
% stan requirements

The model is fitted using the \texttt{stan} function from the
\texttt{rstan} package which needs a special list of data structures
used by the model as well as a model specification expressed in the
Stan language. Every SDT model has some fixed effects structure, since
even when there are no predictors the model parameters can be
expressed as regressed on a vector of ones (i.e., an
intercept). However, many models also have a hierarchical structure
and, if that is the case, this hierarchical structure has to be
specified when using the \texttt{make\_stan\_model} function. This is
done by providing a list of lists of R model formulae. Each list of
model formulae is composed of at least three elements and specifies
correlated random effects of one sampled factor. The element named
\texttt{group} specifies the sampled factor and the elements named
\texttt{delta} and \texttt{gamma} specify which effects are assumed to
vary between the levels of this sampled factor. % Four other optional
% elements can be used to specify the priors on random effects
% distributions. The elements named \texttt{delta\_nu} and
% \texttt{gamma\_nu} correspond to $\nu^{(\bm{\delta})}$ and
% $\nu^{(\bm{\gamma})}$ parameters and the elements named
% \texttt{delta\_scale} and \texttt{gamma\_scale} correspond to
% $\tau^{(\bm{\delta})}$ and $\tau^{(\bm{\delta})}$ parameters.
%
% Of course, random effects must be estimable within the levels of the
% sampled factor and must already be present in the fixed effects
% specification for the respective parameter, at least implicitly.
When \texttt{make\_stan\_model} is used without any arguments it
specifies the SDT model without any random effects. Fixed effects
model matrices are specified by providing a list with at least two
model formulae, named \texttt{delta} and \texttt{gamma}, to the
\texttt{make\_stan\_data} function, described later in this
paper. Non-default priors can be specified by adding optional elements
to the random and fixed effects specification lists, as described in
the \texttt{make\_stan\_data} function documentation.

@@

% KOD Specyfikacja i dopasowanie modelu

@@

\subsubsection{Step 3: fitting the model}
% make_stan_data

@@

\subsubsection{Step 4: assessing the model fit}
% Jak się zbiega

@@

% The reader should be warned that ROC point posterior predictive
% intervals must be interpreted with care, because ROC points are not
% independent and it is in general not correct to expect 95\% coverage
% from 95\% ROC point posterior predictive intervals when the model is
% true.

>>>>

% mixing properties, constrained czy restricted

\documentclass[oneside,a4paper]{article}
\usepackage[sectionbib,bibnewpage]{apacite}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{float}
\usepackage{graphicx}
\usepackage{listings}

@@

\begin{document}
\maketitle
\tableofcontents{}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{plot_sdtr.pdf}
  \caption{Equal variance normal Signal Detection Theory model with
    additional criteria}
  \label{fig:1}
\end{figure}

@@

A not so uncommon practice of analysing data aggregated over subjects,
items or other sampled factors represents an extreme case of ignoring
hierarchical data structure. Invalidity of this approach in the
context of SDT was clearly illustrated with results of simulational
studies by \citeA{morey2008problematic}, although strictly speaking it
requires such demonstrations only in order to make a point about a
specific case. Because SDT is a non-linear model, by definition
expected values of estimates of its parameters based on data
aggregated over sampled factors are not in general true population
averages. Such estimates are biased and inference about the target
population based on such estimates is simply not valid. To give a
concrete example, concider two unbiased participants, one with
$d' = 2$ and one with $d' = 4$. Their expected average accuracy is
given by $(\Phi(1) + \Phi(2)) / 2 = .91$, which corresponds to
$d' = 2.68$, whereas their true average $d'$ is $3$.

@@

In the \texttt{bhsdtr} package sensitivities and criteria are derived
from unconstrained parameters by convenient isomorphisms. This is
easily achieved in case of sensitivity by deriving $d'$ from
$\delta = \ln(d')$ and modelling random effects on $d'$, if any, by
assuming that $\delta$ is normally distributed. Difficulties in
representing the hierarchical structure of the criteria stem from the
fact that even though each criterion considered in isolation is an
unconstrained parameter, the vector of all the criteria is ordered.
We propose to map the $R^{K-1}$ space, representing unconstrained
criteria, to the $K$ dimensional probability simplex space using the
softmax function and to map the simplex space to the ordered criteria
space by means of the inverse normal CDF. Specifically, assume that a
normal distribution is centered on the midpoint between the two
evidence distributions. The value of criterion $i$ is then given by:

@@

The mapping in Eq.~\ref{eq:1} is an isomorphism between the $R^{K-1}$
space and the space of $K-1$ ordered criteria vectors. Elements
$i = 1...K$ of the $\bm{\gamma}$ vector represent relative distances
between adjacent criteria because their exponents represent relative
magnitudes of areas under the normal curve delineated by the pairs of
adjacent criteria:
$e^{\bm{\gamma}_i} / e^{\bm{\gamma}_j} = (\Phi(\bm{c}_i) -
\Phi(\bm{c}_{i-1})) / (\Phi(\bm{c}_j) - \Phi(\bm{c}_{j-1}))$. When
$K=2$ only $\bm{\gamma}_1$ is free to vary and its value directly
represents the direction and magnitude of bias: $\bm{\gamma}_1$ is $0$
when the criterion is placed at the midpoint between the means of
evidence distributions, and the more negative (positive) it is the
more the criterion is placed to the left (right) of midpoint. Our
experience indicates that it is often a good idea to multiply all the
criteria by a value greater than $1$ which is equivalent to making the
mapping distribution in Eq.~\ref{eq:1} wider. This tends to even out
the values of $\bm{\gamma}$, especially when criteria are widely
spread as can happen for moderate to large $d'$ values. This feature
is implemented in the \texttt{bhsdtr} package by introducing a scaling
factor.

@@

@@

@@

\subsubsection{Step 1: preparing the data}

@@

@@

\subsubsection{Step 2: creating the model code}

The model is fitted using the \texttt{stan} function from the
\texttt{rstan} package which needs a special list of data structures
used by the model as well as a model specification expressed in the
Stan language. Every SDT model has some fixed effects structure, since
even when there are no predictors the model parameters can be
expressed as regressed on a vector of ones (i.e., an
intercept). However, many models also have a hierarchical structure
and, if that is the case, this hierarchical structure has to be
specified when using the \texttt{make\_stan\_model} function. This is
done by providing a list of lists of R model formulae. Each list of
model formulae is composed of at least three elements and specifies
correlated random effects of one sampled factor. The element named
\texttt{group} specifies the sampled factor and the elements named
\texttt{delta} and \texttt{gamma} specify which effects are assumed to
vary between the levels of this sampled factor. When
\texttt{make\_stan\_model} is used without any arguments it specifies
the SDT model without any random effects. Fixed effects model matrices
are specified by providing a list with at least two model formulae,
named \texttt{delta} and \texttt{gamma}, to the
\texttt{make\_stan\_data} function, described later in this
paper. Non-default priors can be specified by adding optional elements
to the random and fixed effects specification lists, as described in
the \texttt{make\_stan\_data} function documentation.

@@

@@

\subsubsection{Step 3: fitting the model}

@@

\subsubsection{Step 4: assessing the model fit}

@@
